Optimal updating magnitude in adaptive flat-distribution-sampling simulations

### 2017-10-05 doc/invt.tex
absorb Justin's changes

We will approximate the histogram fluctuation part, $\xi_{*i}(t)$,
%and hence its deviation from the bin average, $\xi_{*i}(t)$,
as white noise such that
%
\begin{equation}
  \langle \xi_{*i}(t) \, \xi_{*i}(t') \rangle
  = G_i \, \delta(t - t')
  ,
  \label{eq:Gi_def}
\end{equation}
%
for some constant $G_i$,
and $\delta(t)$ is Dirac's delta function.
%
We expect the approximation to hold
for simulations much longer than the autocorrelation time.

@@

From Eqs. \eqref{eq:error_def} and
\eqref{eq:vi_solution},
we can write the total error at the end of period $T$
as a sum of a residual error, $E_R(T)$ (from the decay of the initial error)
and an asymptotic error, $E_A(T)$:
%
\begin{align}
  \Err(T)
  =
  \Err_R(T) + \Err_A(T)
  ,
  \label{eq:error_tot}
\end{align}
%
where
\begin{align}
  \Err_R(T)
  &= \Err(0) \, e^{-2 \, q(T)}
  ,
  \label{eq:ER_sbin}
  \\
  \Err_A(T)
  &=
  \sum_{i=1}^n \rho_i
    \int_0^T\!\!\!\int_0^T
    \!\!
    \langle
      \xi_{*i}(t) \xi_{*i}(t')
    \rangle
    \dot \theta\bigl( q(t) \bigr)
    \dot \theta\bigl( q(t') \bigr)
    \, dt dt'
  \notag \\
  &= \Gamma \, \int_0^T \dot \theta^2\bigl( q(t) \bigr) \, dt
  .
  \label{eq:EA_sbin}
\end{align}
In Eq. \eqref{eq:EA_sbin},
we have used Eq. \eqref{eq:Gi_def} and introduced
%$\Gamma \equiv \sum_{i=1}^n \rho_i \, G_i$.
%$\Gamma \equiv \sum_{i=1}^n \rho_i \, G_i
%\approx \sum_{i=1}^n \rho_i \int_{-\infty}^{\infty} \langle \xi_{*i}(t) \xi_{*i}(0) \rangle \, dt$.
\begin{equation}
  \Gamma
  \equiv
  \sum_{i=1}^n \rho_i \, G_i
  \approx
  \sum_{i=1}^n \rho_i \int_0^T
    \langle \xi_{*i}(t) \xi_{*i}(0) \rangle \, dt
  .
  \notag%\label{eq:Gamma_from_G}
\end{equation}

The error depends implicitly on the curve, $\alpha(t)$,
or, equivalently, the curve, $q(t)$.
%via $v_{*i}(T)$.
%
%If $\alpha(t)$ is a constant, $a_0$,
%so that $q(t) = a_0 \, t$,
%then after a long period,
%the error approach an equilibrium value
%%
%\begin{equation}
%  \Err_\mathrm{equil.}
%  = \Err_A = \Gamma \int_0^\infty e^{-2\,q'} \, a_0 \, d q'
%  = \frac{1}{2} a_0 \Gamma
%  .
%  \label{eq:error_equil_sbin}
%\end{equation}
%
We wish to find the $q(t)$
that minimizes this error.
%
It is convenient to
vary the curve, $q(t)$ ($0 < t < T$),
under a fixed value of $q(T)$,
since it allows us to focus on
the asymptotic error, $E_A(T)$.
%
The contribution of the residual error
will be considered later
by adjusting the value of $q(T)$.
%
Note that the asymptotic error
has the form of the action of a free particle
at $\theta\bigl( q(t) \bigr)$,
with the endpoints fixed at
$\theta\bigl( q(0) \bigr)  = e^{- q(T)}$
and
$\theta\bigl( q(T) \bigr) = 1$.
%
Thus, the velocity should be a constant
%Since the residual error
%is fixed during the process,
%the asymptotic error,
%resembling the action of a free particle,
%demands a constant velocity
%
\begin{equation}
  \frac{d}{dt} \theta\bigl( q(t) \bigr) = c
  .
\label{eq:dthetadt_const}
\end{equation}
%
Then, by using Eq. \eqref{eq:theta_def}, we have
%
\begin{equation}
  e^{ q(t) - q(T) }
  =
  c \, (t + t_0)
  =
  \frac{ t + t_0 }
       { T + t_0 }
  ,
\label{eq:expqt}
\end{equation}
%
where $t_0$ is some constant,
and the previous constant, $c$, has been determined
from the $t = T$ case
in the second step.
%
Taking the logarithm,
and differentiating both sides
with respect to $t$,
we get the inverse-time schedule
%
\begin{equation}
  \alpha(t) = \frac{ 1 }{ t + t_0 }
  ,
\label{eq:alpha_invt1}
\end{equation}
%
with Eq. \eqref{eq:alpha_invt} being
the special case of $t_0 = 0$.
%

@@

%To simplify the ensuing discussion,
Second, we limit ourselves to matrices % $\mathbf w$'s
that satisfy
the detailed balance condition for $\pmb\rho$,
%
\begin{equation}
  \rho_i \, w_{ij} = \rho_j \, w_{ji}
  .
  \notag
  %\label{eq:w_detailedbalance}
\end{equation}
%
This requires the scaled updating matrix,
$\sqrt{ \rho_i/\rho_j } \, w_{ij}$,
to be symmetric,
and the symmetry allows the diagonalization
of $\mathbf w$ with a set of
eigenvectors, $\phi_{ki}$,
%
\begin{equation}
  \sum_{i = 1}^n \phi_{ki} \, w_{ij}
  =
  \lambda_k \, \phi_{kj}
  ,
\label{eq:eig_w}
\end{equation}
%
satisfying the orthonormal conditions\cite{vankampen}:
%
\begin{align}
  \sum_{k = 0}^{n - 1}
    \phi_{ki} \, \phi_{kj}
  &=
  \delta_{ij} \, \rho_i,
  \label{eq:eig_orthonormal_cols}
  \\
  \sum_{i = 1}^n
    \frac{ \phi_{ki} \, \phi_{li} }
         { \rho_i }
  &=
  \delta_{kl}
  ,
  \label{eq:eig_orthonormal_rows}
\end{align}
%
where we have started the index of the eigenmodes at $0$
instead of $1$ for later convenience.

@@

We can now use Eq. \eqref{eq:eig_w} to diagonalize
the multiple-bin updating scheme,
Eq. \eqref{eq:mbin_update} [using $u_i(t)$ instead of $v_i(t)$],
as
%
\begin{equation}
  {\tilde v}_k(t + 1) =
  {\tilde v}_k(t) + \alpha(t) \, \lambda_k \,
  {\tilde f}_k(t)
  .
  \label{eq:vkupdate}
\end{equation}
%
In continuous time, we have, for $k > 0$,
%
\begin{equation}
  \dot{\tilde v}_k(t)
  =
  \alpha(t) \, \lambda_k \, {\tilde f}_k(t)
  \approx
  -\alpha(t) \, \lambda_k \,
  \bigl[ {\tilde v}_k(t) - {\tilde \xi}_k(t) \bigr]
  ,
  \label{eq:vt_diffeq_mbin}
\end{equation}
%
where
we have Fourier transformed Eq. \eqref{eq:sh_ave}
and used Eq. \eqref{eq:xtstar}
in the second step.

@@

By Eq. \eqref{eq:parseval},
we can rewrite the error defined in Eq. \eqref{eq:error_def} as
a sum of the errors in individual modes\footnote{If
the error weighting distribution differs from $\rho_i$
in Eq. \eqref{eq:error_def},
the error would contain quadratic cross terms among different eigenmodes.
%
While the solution for the optimal schedule retains the same form,
the mass function defined in Eq. \eqref{eq:mass_func}
will include the corresponding cross terms.
}
%$\Err(T) = \sum_{k = 1}^{n - 1} \left\langle {\tilde v}_k^2(T) \right\rangle$.
%
\begin{align}
  \Err(T)
  =
  \sum_{k = 1}^{n - 1}
    \left\langle
      {\tilde v}_k^2(T)
    \right\rangle
  .
  \notag
\end{align}
%
As in Eq. \eqref{eq:error_tot},
we can decompose the error as the sum
of the residual error and the asymptotic error, with
\begin{align}
  \Err_R(T)
  &=
  \sum_{k = 1}^{n-1}
    \left\langle
      {\tilde v}_k^2(0)
    \right\rangle \,
    e^{ - 2 \, \lambda_k  \, q(T) }
  ,
  \label{eq:error_res}
  \\
  \Err_A(T)
  &=
  \sum_{k=1}^{n-1}
  \int_0^T\!\!\!\int_0^T
  \!\!
  \langle
    \tilde\xi_k(t)
    \tilde\xi_k(t')
  \rangle
  \theta_k\bigl(q(t)\bigr)
  \theta_k\bigl(q(t')\bigr)
  \, dtdt'
  \notag \\
  &=
  \int_0^T
  \sum_{k = 1}^{n-1}
  \Gamma_k \, \dot \theta_k^2\bigl( q(t) \bigr) \, dt
  ,
  \label{eq:error_asym}
\end{align}
%
where
%
%$\Gamma_k$ is computed from
%the autocorrelation function
%of the histogram fluctuation,
%$\bigl\langle {\tilde \xi}_k(0)
%\, {\tilde \xi}_k(t) \bigr\rangle$,
%as:
%
\begin{equation}
  \Gamma_k
  \approx
  \int_{-\infty}^\infty
  \bigl\langle
    \tilde\xi_k(t) \, \tilde\xi_k(0)
  \bigr\rangle
  \, dt
  \to
  \sum_{t = -\infty}^\infty
  \bigl\langle
    \tilde\xi_k(t) \, \tilde\xi_k(0)
  \bigr\rangle
  ,
  \label{eq:Gamma_sum}
\end{equation}
%
with the second expression being
the counterpart in the discrete-time setting.
%
Equation \eqref{eq:Gamma_sum} represents
a first approximation for modeling
a general fluctuation, ${\tilde \xi}_k(t)$,
as an equivalent white noise.
%
In this way,
the underlying sampling process
affects the error only through the
few numbers, $\Gamma_k$'s.
%
Note that by Eq. \eqref{eq:parseval}, we have
$\sum_{k=1}^{n-1} \bigl\langle {\tilde \xi}_k(t) \, {\tilde \xi}_k(0) \bigr\rangle
= \sum_{i=1}^n \rho_i \bigl\langle \xi_{*i}(t) \, \xi_{*i}(0) \bigr\rangle,$
and summing $t$ from $-\infty$ to $\infty$ yields,
\begin{equation}
  \sum_{k=1}^{n-1} \Gamma_k = \sum_{i=1}^n \rho_i \, G_i = \Gamma
  .
  \label{eq:Gammak_sum}
\end{equation}
This shows the equivalence of
Eq. \eqref{eq:EA_sbin} and \eqref{eq:error_asym}
in the case of the single-bin updating scheme, where $\lambda_k \equiv 1$.
%In Appendix \ref{sec:Gamma_measure},
%we discuss a way of estimating $\Gamma_k$
%from an FDS simulation under a constant magnitude.
%

@@

As in the single-bin case,
by varying the schedule, $\alpha(t)$, under a fixed value of
$q(T) = \int_0^T \alpha(t) \, dt$,
we can focus on the asymptotic error,
$\Err_A(T)$.
%
The value of $q(T)$ will be adjusted later
to minimize the total error. % in Sec. \ref{sec:optinitalpha}.
%
We can rewrite Eq. \eqref{eq:error_asym} much like an action
of a particle whose position is given by $q(t)$:
%
\begin{equation}
  \Err_A(T)
  =
  \int_0^T
    {\mathcal L} \bigl[ q(t)\bigr]
    \, dt
  ,
  \notag
  %\label{eq:error_asym_Lagrangian}
\end{equation}
%
where the Lagrangian is
%
\begin{align}
  {\mathcal L} \bigl[ q(t) \bigr]
  &=
  \sum_{ k = 1 }^{n - 1}
    \Gamma_k \, {\dot \theta}_k^2\bigl[ q(t) \bigr]
  %\notag
  %\\
  %&=
  =
  %\sum_{ k = 1 }^{n - 1}
  %  \Gamma_k \, \lambda_k^2 \, \theta_k^2 \bigl[ q(t) \bigr]
  %\; \dot q^2( t )
  %\notag
  %\\
  %&=
  M^2\bigl(q(T) - q(t) \bigr)
  \; \dot q^2( t )
  .
\notag
\end{align}
%
where we have
used Eq. \eqref{eq:uk_def} and
defined the (square-root) mass function as
%
\begin{equation}
  M(\bar q)
  \equiv
  \sqrt{
    \textstyle\sum_{ k = 1 }^{n - 1}
    \Gamma_k \, \lambda_k^2 \, e^{-2 \, \lambda_k \, \bar q}
  }
  .
  %\notag
  \label{eq:mass_func}
\end{equation}
%
In this mechanical analogy,
the schedule, $\dot q(t) = \alpha(t)$,
corresponds to the velocity of the particle,
%
and the above Lagrangian, containing only the kinetic energy,
describes a free particle
with a position-dependent mass.
%
Here, the Hamiltonian coincides with the Lagrangian:
%
\begin{equation}
  \mathcal H
  =
  \frac{ \partial \mathcal L }
       { \partial \dot q     }
  \, \dot q
  -
  \mathcal L
  =
  2 \, \mathcal L
  - \mathcal L
  =
  \mathcal L
  .
  \notag
  %\label{eq:error_asym_Hamiltonian}
\end{equation}
%
Since the Lagrangian
does not explicitly depend on time, $t$,
the Hamiltonian is conserved,\cite{goldstein, *landau_mechanics, *arnold, *jose}
i.e. the asymptotic error grows
linearly with time at a rate of
$\mathcal L = \mathcal H$.
%
We may set
%
\begin{equation}
  \sqrt{ \mathcal H }
  =
  \sqrt{ \mathcal L }
  =
  M\bigl( q(T) - q(t) \bigr)
  \;
  \dot q(t)
  =
  \frac{C_M}{T}
  ,
  \label{eq:Lagrangian_const}
\end{equation}
%
for some positive $C_M$.
%
Equation \eqref{eq:Lagrangian_const} serves
as a generalization of Eq. \eqref{eq:dthetadt_const}.
%
By integrating this equation of motion, we get
%
\begin{equation}
  \int_{ q(T) - q(t) }^{ q(T) }
    M(\bar q)
    \;
    d \bar q
  =
  C_M \, \frac t T
  ,
  \label{eq:q_opt}
\end{equation}
%
and $C_M$ can be determined from
the $t = T$ case as
%
\begin{equation}
  C_M =
  \int_{ 0 }^{ q(T) }
    M( \bar q )
    \;
    d \bar q
  .
  \label{eq:mint}
\end{equation}
%
Thus, Eq. \eqref{eq:Lagrangian_const}
gives an implicit equation for $q(t)$,
and hence the optimal schedule,
$\alpha(t) = \dot q(t)$.
%
This solution can be characterized geometrically,
as shown in Appendix \ref{sec:schedule_geometry}.
%
The minimal asymptotic error is
%
\begin{equation}
  \Err_A(T)
  =
  \mathcal L \, T
  =
  \frac { C_M^2 } { T }
  =
  \frac 1 T
  \left(
    \int_0^{ q(T) } M(\bar q) \, d \bar q
  \right)^{\!\!2}
  .
\label{eq:error_asym2}
\end{equation}

@@

\item \label{step:alpha}
Compute the optimal schedule by
numerically integrating
Eqs. \eqref{eq:q_opt} and \eqref{eq:mint}.\footnote{We
  used the trapezoidal rule for the integration.
  The grid points were
  $q_{[i]} = \bigl(q(T) + 1\bigr) - \bigl(q(T)+1\bigr)^{1-i/N_g}$,
  where $i = 0, \dots, N_g$
  with $N_g = 20000$.
  %
  The grid points were designed for long simulations
  under the Gaussian updating scheme:
  for a mass function, $M(\bar q)$, that is roughly
  inversely proportional to $\bar q$
  (cf. Appendix \ref{sec:schedule_geometry}),
  this grid makes the integral,
  $t_{[i+1]} - t_{[i]}$,
  approximately a constant.
  %
  The resulting values, $\bigl(q_{[i]}, t_{[i]} \bigr)$,
  were used to compute the updating magnitude
  as (cf. Appendix \ref{sec:ave_ub}):
  $$%\begin{equation}
    \alpha_{[i]}
    =
    \frac{1}{\lambda_1}
    \left[
      1
      -
      \exp\left(
      -\lambda_1
      \frac{ q_{[i+1]} - q_{[i-1]} }
           { t_{[i+1]} - t_{[i-1]} }
      \right)
    \right]
    ,
    %\notag
    %\label{eq:alphai}
  $$%\end{equation}
  with
  $q_{[ -1]} \equiv t_{[-1]} \equiv 0$,
  $q_{[N_g+1]} \equiv q(T)$,
  and
  $t_{[N_g+1]} \equiv T$.
  %
  The values of $(t_{[i]}, \alpha_{[i]})$
  were linearly interpolated
  to get the updating magnitude, $\alpha(t)$, at any $t$.
  }
%

@@

For example,
to sample a flat distribution, $\rho(z) = 1/\Delta z$,
over $\left(z_c - \frac 1 2 \Delta z, z_c + \frac 1 2 \Delta z\right)$,
we can use the Legendre polynomials\cite{arfken}.
%
If the interval is sufficiently small,
we can use $K = 2$ to limit the basis to the first two modes,
$R_0(z) = 1$ and $R_1(z) = \sqrt{12} \, (z - z_c)/\Delta z$
(with $\sigma_z = \Delta z/\sqrt{12}$).
%
%The bias potential is a linear function of $z$,
If we drop the $k = 0$ mode
that corresponds to a uniform shift of the bias potential,
the updating scheme is essentially the LLR algorithm\cite{langfeld2012}
in the single-interval case,
and the inverse-time schedule
was recommended for the optimal convergence of $c_1$.\cite{pellegrini2014}

>>>>

We will approximate the histogram fluctuation part, $\xi_{*i}(t)$,
%and hence its deviation from the bin average, $\xi_{*i}(t)$,
as white noise such that
%
\begin{equation}
  \bigl\langle \xi_{*i}(t) \, \xi_{*i}(t') \bigr\rangle
  = G_i \, \delta(t - t')
  ,
  \label{eq:Gi_def}
\end{equation}
%
for some constant $G_i$,
and $\delta(t)$ is Dirac's delta function.
%
We expect the approximation to hold
for simulations much longer than the autocorrelation time.

@@

From Eqs. \eqref{eq:error_def} and
\eqref{eq:vi_solution},
we can write the total error at the end of period $T$
as a sum of a residual error, $E_R(T)$ (from the decay of the initial error)
and an asymptotic error, $E_A(T)$:
%
\begin{align}
  \Err(T)
  =
  \Err_R(T) + \Err_A(T)
  ,
  \label{eq:error_tot}
\end{align}
%
where
\begin{align}
  \Err_R(T)
  &= \Err(0) \, e^{-2 \, q(T)}
  ,
  \label{eq:ER_sbin}
  \\
  \Err_A(T)
  &=
  \sum_{i=1}^n \rho_i
    \int_0^T\!\!\!\int_0^T
    \!\!
    \bigl\langle
      \xi_{*i}(t) \xi_{*i}(t')
    \bigr\rangle
    \,
    \dot \theta\bigl( q(t) \bigr)
    \dot \theta\bigl( q(t') \bigr)
    \, dt \, dt'
  \notag \\
  &= \Gamma \, \int_0^T \dot \theta^2\bigl( q(t) \bigr) \, dt
  .
  \label{eq:EA_sbin}
\end{align}
In Eq. \eqref{eq:EA_sbin},
we have used Eq. \eqref{eq:Gi_def} and introduced
%$\Gamma \equiv \sum_{i=1}^n \rho_i \, G_i$.
%$\Gamma \equiv \sum_{i=1}^n \rho_i \, G_i
%\approx \sum_{i=1}^n \rho_i \int_{-\infty}^{\infty} \langle \xi_{*i}(t) \xi_{*i}(0) \rangle \, dt$.
\begin{equation}
  \Gamma
  \equiv
  \sum_{i=1}^n \rho_i \, G_i
  \approx
  \sum_{i=1}^n \rho_i \int_0^T
    \bigl\langle \xi_{*i}(t) \xi_{*i}(0) \bigr\rangle
    \, dt
  .
  \notag%\label{eq:Gamma_from_G}
\end{equation}

The error depends implicitly on the curve, $\alpha(t)$,
or, equivalently, the curve, $q(t)$.
%via $v_{*i}(T)$.
%
%If $\alpha(t)$ is a constant, $a_0$,
%so that $q(t) = a_0 \, t$,
%then after a long period,
%the error approach an equilibrium value
%%
%\begin{equation}
%  \Err_\mathrm{equil.}
%  = \Err_A = \Gamma \int_0^\infty e^{-2\,q'} \, a_0 \, d q'
%  = \frac{1}{2} a_0 \Gamma
%  .
%  \label{eq:error_equil_sbin}
%\end{equation}
%
We wish to find the $q(t)$
that minimizes this error.
%
It is convenient to
vary the curve, $q(t)$ ($0 < t < T$),
under a fixed value of $q(T)$,
since it allows us to focus on
the asymptotic error, $E_A(T)$.
%
The contribution of the residual error
will be considered later
by adjusting the value of $q(T)$.
%
Note that the asymptotic error
has the form of the action of a free particle
at $\theta\bigl( q(t) \bigr)$,
with the endpoints fixed at
$\theta\bigl( q(0) \bigr)  = e^{- q(T)}$
and
$\theta\bigl( q(T) \bigr) = 1$.
%
Thus, the velocity should be a constant
%Since the residual error
%is fixed during the process,
%the asymptotic error,
%resembling the action of a free particle,
%demands a constant velocity
%
\begin{equation}
  \frac{d}{dt} \theta\bigl( q(t) \bigr) = c
  .
\label{eq:dthetadt_const}
\end{equation}
%
Then, by using Eq. \eqref{eq:theta_def}, we have
%
\begin{equation}
  e^{ q(t) - q(T) }
  =
  c \, (t + t_0)
  =
  \frac{ t + t_0 }
       { T + t_0 }
  ,
\label{eq:expqt}
\end{equation}
%
where $t_0$ is some constant,
and the previous constant, $c$, has been determined
from the $t = T$ case
in the second step.
%
Taking the logarithm,
and differentiating both sides
with respect to $t$,
we get the inverse-time schedule
%
\begin{equation}
  \alpha(t) = \frac{ 1 }{ t + t_0 }
  ,
\label{eq:alpha_invt1}
\end{equation}
%
and Eq. \eqref{eq:alpha_invt} is
the special case of $t_0 = 0$.
%

@@

%To simplify the ensuing discussion,
Second, we limit ourselves to matrices % $\mathbf w$'s
that satisfy
the detailed balance condition for $\pmb\rho$,
%
\begin{equation}
  \rho_i \, w_{ij} = \rho_j \, w_{ji}
  .
  \notag
  %\label{eq:w_detailedbalance}
\end{equation}
%
This requires the scaled updating matrix,
$\sqrt{ \rho_i/\rho_j } \, w_{ij}$,
to be symmetric,
and the symmetry allows the diagonalization
of $\mathbf w$ with a set of
eigenvectors, $\phi_{ki}$, as
%
\begin{equation}
  \sum_{i = 1}^n \phi_{ki} \, w_{ij}
  =
  \lambda_k \, \phi_{kj}
  ,
\label{eq:eig_w}
\end{equation}
%
with the $\phi_{ki}$'s satisfying the orthonormal conditions\cite{vankampen}:
%
\begin{align}
  \sum_{k = 0}^{n - 1}
    \phi_{ki} \, \phi_{kj}
  &=
  \delta_{ij} \, \rho_i,
  \label{eq:eig_orthonormal_cols}
  \\
  \sum_{i = 1}^n
    \frac{ \phi_{ki} \, \phi_{li} }
         { \rho_i }
  &=
  \delta_{kl}
  .
  \label{eq:eig_orthonormal_rows}
\end{align}
%
Here we have started the index of the eigenmodes at $0$
instead of $1$ for later convenience.

@@

We can now use Eq. \eqref{eq:eig_w} to diagonalize
the multiple-bin updating scheme,
Eq. \eqref{eq:mbin_update}
[using $v_i(t)$ instead of $u_i(t)$],
as
%
\begin{equation}
  {\tilde v}_k(t + 1) =
  {\tilde v}_k(t) + \alpha(t) \, \lambda_k \,
  {\tilde f}_k(t)
  .
  \label{eq:vkupdate}
\end{equation}
%
In continuous time, we have, for $k > 0$,
%
\begin{equation}
  \dot{\tilde v}_k(t)
  =
  \alpha(t) \, \lambda_k \, {\tilde f}_k(t)
  \approx
  -\alpha(t) \, \lambda_k \,
  \bigl[ {\tilde v}_k(t) - {\tilde \xi}_k(t) \bigr]
  ,
  \label{eq:vt_diffeq_mbin}
\end{equation}
%
where
we have Fourier transformed Eq. \eqref{eq:sh_ave}
and used Eq. \eqref{eq:xtstar}
in the second step.

@@

By Eq. \eqref{eq:parseval},
we can rewrite the error defined in Eq. \eqref{eq:error_def} as
a sum of the errors in individual modes\footnote{If
the error weighting distribution differs from $\rho_i$
in Eq. \eqref{eq:error_def},
the error would contain quadratic cross terms among different eigenmodes.
%
While the solution for the optimal schedule retains the same form,
the mass function defined in Eq. \eqref{eq:mass_func}
will include the corresponding cross terms.
}
%$\Err(T) = \sum_{k = 1}^{n - 1} \left\langle {\tilde v}_k^2(T) \right\rangle$.
%
\begin{align}
  \Err(T)
  =
  \sum_{k = 1}^{n - 1}
    \left\langle
      {\tilde v}_k^2(T)
    \right\rangle
  .
  \notag
\end{align}
%
As in Eq. \eqref{eq:error_tot},
we can decompose the error as the sum
of the residual error and the asymptotic error, with
\begin{align}
  \Err_R(T)
  &=
  \sum_{k = 1}^{n-1}
    \left\langle
      {\tilde v}_k^2(0)
    \right\rangle \,
    e^{ - 2 \, \lambda_k  \, q(T) }
  ,
  \label{eq:error_res}
  \\
  \Err_A(T)
  &=
  \sum_{k=1}^{n-1}
  \int_0^T\!\!\!\int_0^T
  \!\!
  \bigl\langle
    \tilde\xi_k(t)
    \tilde\xi_k(t')
  \bigr\rangle \,
  \theta_k\bigl(q(t)\bigr)
  \theta_k\bigl(q(t')\bigr)
  \, dt \, dt'
  \notag \\
  &=
  \int_0^T
  \sum_{k = 1}^{n-1}
  \Gamma_k \, \dot \theta_k^2\bigl( q(t) \bigr) \, dt
  ,
  \label{eq:error_asym}
\end{align}
%
where
%
%$\Gamma_k$ is computed from
%the autocorrelation function
%of the histogram fluctuation,
%$\bigl\langle {\tilde \xi}_k(0)
%\, {\tilde \xi}_k(t) \bigr\rangle$,
%as:
%
\begin{equation}
  \Gamma_k
  \approx
  \int_{-\infty}^\infty
  \bigl\langle
    \tilde\xi_k(t) \, \tilde\xi_k(0)
  \bigr\rangle
  \, dt
  \to
  \sum_{t = -\infty}^\infty
  \bigl\langle
    \tilde\xi_k(t) \, \tilde\xi_k(0)
  \bigr\rangle
  ,
  \label{eq:Gamma_sum}
\end{equation}
%
with the second expression being
the counterpart in the discrete-time setting.
%
Equation \eqref{eq:Gamma_sum} represents
a first approximation for modeling
a general fluctuation, ${\tilde \xi}_k(t)$,
as an equivalent white noise.
%
In this way,
the underlying sampling process
affects the error only through the
few numbers, $\Gamma_k$'s.
%
Note that by Eq. \eqref{eq:parseval}, we have
$\sum_{k=1}^{n-1} \bigl\langle {\tilde \xi}_k(t) \, {\tilde \xi}_k(0) \bigr\rangle
= \sum_{i=1}^n \rho_i \bigl\langle \xi_{*i}(t) \, \xi_{*i}(0) \bigr\rangle,$
and summing $t$ from $-\infty$ to $\infty$ yields,
\begin{equation}
  \sum_{k=1}^{n-1} \Gamma_k = \sum_{i=1}^n \rho_i \, G_i = \Gamma
  .
  \label{eq:Gammak_sum}
\end{equation}
This shows the equivalence of
Eq. \eqref{eq:EA_sbin} and \eqref{eq:error_asym}
in the case of the single-bin updating scheme, where $\lambda_k \equiv 1$.
%In Appendix \ref{sec:Gamma_measure},
%we discuss a way of estimating $\Gamma_k$
%from an FDS simulation under a constant magnitude.
%

@@

As in the single-bin case,
by varying the schedule, $\alpha(t)$, under a fixed value of
$q(T) = \int_0^T \alpha(t) \, dt$,
we can focus on the asymptotic error,
$\Err_A(T)$.
%
The value of $q(T)$ will be adjusted later
to minimize the total error. % in Sec. \ref{sec:optinitalpha}.
%
We can rewrite Eq. \eqref{eq:error_asym} much like the action
of a particle whose position is given by $q(t)$:
%
\begin{equation}
  \Err_A(T)
  =
  \int_0^T
    {\mathcal L} \bigl[ q(t)\bigr]
    \, dt
  ,
  \notag
  %\label{eq:error_asym_Lagrangian}
\end{equation}
%
where the Lagrangian is
%
\begin{align}
  {\mathcal L} \bigl[ q(t) \bigr]
  &=
  \sum_{ k = 1 }^{n - 1}
    \Gamma_k \, {\dot \theta}_k^2\bigl[ q(t) \bigr]
  %\notag
  %\\
  %&=
  =
  %\sum_{ k = 1 }^{n - 1}
  %  \Gamma_k \, \lambda_k^2 \, \theta_k^2 \bigl[ q(t) \bigr]
  %\; \dot q^2( t )
  %\notag
  %\\
  %&=
  M^2\bigl(q(T) - q(t) \bigr)
  \; \dot q^2( t )
  .
\notag
\end{align}
%
where we have
used Eq. \eqref{eq:uk_def} and
defined the (square-root) mass function as
%
\begin{equation}
  M(\bar q)
  \equiv
  \sqrt{
    \textstyle\sum_{ k = 1 }^{n - 1}
    \Gamma_k \, \lambda_k^2 \, e^{-2 \, \lambda_k \, \bar q}
  }
  .
  %\notag
  \label{eq:mass_func}
\end{equation}
%
In this mechanical analogy,
the schedule, $\dot q(t) = \alpha(t)$,
corresponds to the velocity of the particle,
%
and the above Lagrangian, containing only the kinetic energy,
describes a free particle
with a position-dependent mass.
%
Here, the Hamiltonian coincides with the Lagrangian:
%
\begin{equation}
  \mathcal H
  =
  \frac{ \partial \mathcal L }
       { \partial \dot q     }
  \, \dot q
  -
  \mathcal L
  =
  2 \, \mathcal L
  - \mathcal L
  =
  \mathcal L
  .
  \notag
  %\label{eq:error_asym_Hamiltonian}
\end{equation}
%
Since the Lagrangian
does not explicitly depend on time, $t$,
the Hamiltonian is conserved,\cite{goldstein, *landau_mechanics, *arnold, *jose}
i.e. the asymptotic error grows
linearly with time at a rate of
$\mathcal L = \mathcal H$.
%
We may set
%
\begin{equation}
  \sqrt{ \mathcal H }
  =
  \sqrt{ \mathcal L }
  =
  M\bigl( q(T) - q(t) \bigr)
  \;
  \dot q(t)
  =
  \frac{C_M}{T}
  ,
  \label{eq:Lagrangian_const}
\end{equation}
%
for some positive $C_M$.
%
Equation \eqref{eq:Lagrangian_const} serves
as a generalization of Eq. \eqref{eq:dthetadt_const}.
%
By integrating this equation of motion, we get
%
\begin{equation}
  \int_{ q(T) - q(t) }^{ q(T) }
    M(\bar q)
    \;
    d \bar q
  =
  C_M \, \frac t T
  ,
  \label{eq:q_opt}
\end{equation}
%
and $C_M$ can be determined from
the $t = T$ case as
%
\begin{equation}
  C_M =
  \int_{ 0 }^{ q(T) }
    M( \bar q )
    \;
    d \bar q
  .
  \label{eq:mint}
\end{equation}
%
Thus, Eq. \eqref{eq:Lagrangian_const}
gives an implicit equation for $q(t)$,
and hence the optimal schedule,
$\alpha(t) = \dot q(t)$.
%
This solution can be characterized geometrically,
as shown in Appendix \ref{sec:schedule_geometry}.
%
The minimal asymptotic error is
%
\begin{equation}
  \Err_A(T)
  =
  \mathcal L \, T
  =
  \frac { C_M^2 } { T }
  =
  \frac 1 T
  \left(
    \int_0^{ q(T) } M(\bar q) \, d \bar q
  \right)^{\!\!2}
  .
\label{eq:error_asym2}
\end{equation}

@@

\item \label{step:alpha}
Compute the optimal schedule by
numerically integrating
Eqs. \eqref{eq:q_opt} and \eqref{eq:mint}.\footnote{We
  used the trapezoidal rule for the integration.
  The grid points were
  $q_{[i]} = \bigl(q(T) + 1\bigr) - \bigl(q(T)+1\bigr)^{1-i/N_g}$,
  where $i = 0, \dots, N_g$
  with $N_g = 20000$.
  %
  The grid points were designed for long simulations
  under the Gaussian updating scheme:
  for a mass function, $M(\bar q)$,
  that is roughly proportional to $1/\bar q$
  (cf. Appendix \ref{sec:schedule_geometry}),
  this grid makes the integral,
  $t_{[i+1]} - t_{[i]}$,
  approximately a constant.
  %
  The resulting values, $\bigl(q_{[i]}, t_{[i]} \bigr)$,
  were used to compute the updating magnitude
  as (cf. Appendix \ref{sec:ave_ub}):
  $$%\begin{equation}
    \alpha_{[i]}
    =
    \frac{1}{\lambda_1}
    \left[
      1
      -
      \exp\left(
      -\lambda_1
      \frac{ q_{[i+1]} - q_{[i-1]} }
           { t_{[i+1]} - t_{[i-1]} }
      \right)
    \right]
    ,
    %\notag
    %\label{eq:alphai}
  $$%\end{equation}
  with
  $q_{[ -1]} \equiv t_{[-1]} \equiv 0$,
  $q_{[N_g+1]} \equiv q(T)$,
  and
  $t_{[N_g+1]} \equiv T$.
  %
  The values of $(t_{[i]}, \alpha_{[i]})$
  were linearly interpolated
  to get the updating magnitude, $\alpha(t)$, at any $t$.
  }
%

@@

For example,
to sample a flat distribution, $\rho(z) = 1/\Delta z$,
over $\left(z_c - \frac 1 2 \Delta z, z_c + \frac 1 2 \Delta z\right)$,
we can use the Legendre polynomials\cite{arfken}.
%
If the interval is sufficiently small,
we can use $K = 2$ to limit the basis to the first two modes,
$R_0(z) = 1$ and $R_1(z) = \sqrt{12} \, (z - z_c)/\Delta z$
(with $\sigma_z = \Delta z/\sqrt{12}$).
%
%The bias potential is a linear function of $z$,
If we drop the $k = 0$ mode
that corresponds to a uniform shift of the bias potential,
the updating scheme is essentially the LLR algorithm\cite{langfeld2012}
in the single-interval case,
and the inverse-time schedule
has been recommended for the optimal convergence of $c_1$.\cite{pellegrini2014}

In practice,
the above algorithm is often applied simultaneously
to an array of adjacent intervals
to cover a wide range of the collective variable, $z$.\cite{langfeld2012, pellegrini2014}
%
Note that in this expanded ensemble setup,
the updating equation for $c_0$ is no longer trivial,
but corresponds to a single-bin updating scheme of
constructing the bias potential for transitions among different intervals.
%
The WL algorithm is in fact recovered by the $K = 1$ case.
%
However, the LLR algorithm estimates $c_0$ from
an approximate relation similar to Eq. \eqref{eq:c0_est} below.
%
%In this way, attempts to move out of the current interval
%are not necessarily rejected,
%but can be interpreted as transitions to
%another internal and accepted with proper probability.
